Dark matter web in the Universe

### 2018-06-06 Draft/Chapter6/concl.tex
thesis update

\chapter{Conclusions and Future}

In this Chapter, we present our general conclusions. We discuss the limitations inherent
in our model, and speculate upon how we may address these weaknesses in the future.
We, furthermore, speculate upon the possible applications of our model to real plasmas.
We, finally, consider the physical questions still remaining – concluding, ultimately, that a
considerable amount of work is still needed to answer these questions definitively.

@@

\subsection{Flip-Flop}
\subsection{Caustics}
\subsection{Simulation}

@@

>>>>

\chapter{General Conclusions}\label{chapter6}

We have reported the results of the first studies of the multi-stream environment in cosmological context - from dark matter halo environments to connectivity in single streaming voids.

\section{Summary of major results}

The full dynamical state of  dark matter can be described as a three-dimensional sub-manifold
in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
(where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis. Counting the number of velocity streams in gravitational collapses supplements our knowledge of spatial clustering. Our major results can be summarized as follows.

\subsection{Heirarchial structure of the dark matter web}

The first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations
in the $\Lambda$CDM cosmology was reported in \cite{Ramachandra2015}. At the resolution of the simulation i.e. without additional smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly  at the filament crossings, and each filament is embedded in the wall like fabric of the web at the wall crossings. Locally, each halo or sub-halo is a peak in the number of streams field. The number of streams in the neighbouring filaments is higher than in the neighbouring walls. The walls are regions where number of streams is equal to three or a few. Voids are uniquely defined by the local condition requiring to be a single-stream flow region. We demonstrated in Chapter \ref{chapter3} that the shells of streams around haloes are quite thin and the closest void region is typically within one and a half FOF radius from the center of the halo.

\subsection{Dark matter haloes}

In our simulation reported in Chapter \ref{chapter3}, simple virial density criteria corresponds to a global threshold of $n_{str} \geq 90$ for DM haloes. This shows a reasonably good correspondence with structure finders. A more sophisticated halo finding formulation is done in Chapter \ref{chapter5} assuming that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field. In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites. This halo finding algorithm is free of any heuristic parameters, and maybe extended to delineate filaments, walls, voids, as well as halo sub-structures.

\subsection{Topological transitions in the web}

Topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields, as shown in Chapter \ref{chapter4}. A single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than simulation resolution reveal tiny pockets of voids that are isolated by membranes of the structure. For the multistreaming excursion sets, the percolating structure is significantly  thinner than the filaments in over-density excursion approach.

@@

Although the scope of this thesis has primarily been multistream field and its application to cosmic structure formation, it is necessary to highlight other important physical quantities derived from the Lagrangian submanifold. For instance, one may track the {\it Flip-Flop} field - a to understand the rich and comples substructures within haloes. On the other hand, Lagrangian submanifold may also be used to identify caustic surfaces -- measure-zero surfaces in the DM Universe that have formally infinite density.

This section reviews three potential avenues for interesting analyses on Lagrangian submanifold.

\subsection{Flip-Flop}

The Flip-Flop field, $n_{ff}(\mathbf{q})$ is defined in Lagrangian space by the number of turns experienced by each DM fluid particle. This number is estimated in three dimensions by computing the Jacobian  $J(\mathbf{q}, t) = |\frac{\partial\mathbf{x}}{\partial\mathbf{q}}|$  on each particle at each time step. If the sign of the Jacobian changes, the number of flip-flops for the corresponding particles is increased by one. Details of calculation of flip-flop field and it's significance in understand dynamical history of substructure formation

\subsection{Caustics}

\subsubsection{Geometric characterization of Caustics}

The inter-particle separation between DM particles in Eulerian space is a generally a good measure of clustering. DM haloes detected by Friends-of-Friends algorithm, for example, are expected to be to have a separation less than 0.2 $\times$ the mean separation of particles in the simulation. Similarly other heuristic thresholds for walls and filaments can be calculated too.

The DM particles on the caustics are on curved 2-Dimensional surfaces, forming vertices of triangles that constitute the {\it tiles} of the surface. These triangles are of various shapes and size. Since the mean separation of particles is smaller than elsewhere, the triangles are smaller on an average. Figure \ref{fig:PDF_lmax} shows this variation of area of the triangles. Over 8 percent of the triangles have areas less than $10^{-6} h^{-2}Mpc^2$. The distribution function reduces roughly as a power law with the square root of area.

\begin{figure} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/PDF_triangle.pdf} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/Scatter_Al_ff.png} 
\caption{Top: PDF of triangle discriptors: $l_{max}$ is the largest side of triangles in the caustic surfaces, $h$ is the height, and $\sqrt{A}$ is the square root of area of each triangle. Bottom: Scatter plot of $l_{max}$ and $h$, with $\sqrt{A}$ along the colorbar. }
\label{fig:PDF_lmax}
\end{figure}

However, it was visually seen that several non-compact triangles with small areas were selected. A better heuristic parameter was found to be the longest side of the triangle, $l_{max}$. Choosing this parameter, we circumnavigate the issue of selecting elongated triangles. Variation of $l_{max}$ \hl{is different than that of square root of area} (Figure \ref{fig:PDF_lmax}.)

Finally, using the heuristic parameter of $l_{max}$, we are able to isolate caustic surfaces around specific structures like the haloes. The Figure \ref{fig:caustic_lmax} shows all the caustic surfaces in the top left column, and the inner caustics filtered using $l_{max}$ thresholds in the subsequent coulmns. Furthermore, compared to Figure \ref{fig:caustic_ff}, the inner caustics visually seem to be less elongated. Another interesting aspect of this filtering scheme is the ability to isolate the caustic surfaces around the haloes as seen in the bottom right panel of Figure \ref{fig:caustic_lmax}.

\begin{figure*} 
\centering\includegraphics[height=15cm]{Chapter2/Plots/caustic_connectivity.png} 
\caption{$L=20$ Mpc, $N_p = 32^3$. Small simulation, Caustic are filtered by longest side length of the triangle: Top left: $ l_{max} \geq 0$, Top right: $ l_{max} \geq 1$, Bottom left: $ l_{max} \geq 2$, Bottom right: $ l_{max} \geq 3$. }
\label{fig:caustic_lmax}
\end{figure*}

% Combine - caustics and flip-flop to get levels of caustics! However there might be discreetness effects, i.e., say a triangle identified as caustic may have edge particles with 0, 1, 2, 3 .. flip-flops (?).

% Check this in a small simulation - Run About L = 200/150 Mpc and Np = $32^3$ and pick up highest stream regions - shouldn't be too much. And then see if you can run a zoom-in simulation around halo-site.

\begin{comment}
\begin{figure} 
\centering\includegraphics[height=15cm]{Plots/percolation.pdf} 
\caption{$L=100$ Mpc (? confirm), $N_p = 128^3$. From Prof. Shandarin - percolation of caustic surface - triangle connectivity is checked. }
\label{fig:1d}
\end{figure}
\end{comment}

Careful examination of the smaller caustic surfaces around the haloes may reveal concentric shells where the particles {\it turn back}. A spherical approximation of this, called the {\it splashback radius} has been analysed by \cite{More2015}. This caustic finding scheme with appropriate filtering is provides a general method to find these boundaries exactly.

\begin{figure} 
\centering\includegraphics[width=9cm]{Chapter2/Plots/f1fES_caustic.pdf} 
\caption{Percolation in caustic particles with variation of $l_{max}$: Top panel shows the mass fraction of the largest isolated caustic surface $f_1$ and mass fraction of all particles in the caustic $f_{ES}$. Bottom panel shows the filling fraction $f_1/f_{ES}$. Between 3.5-4 $h^{-1} Mpc$ the caustic surfaces transition into distinct isolated surfaces -- i.e., forming turn-around boundaries of haloes. }
\label{fig:f1fES}
\end{figure}

% \subsection{Other avenues}
%  Splashback radii, Merger trees, improvements in simulation etc.

@@

%% From Paper2017a intro

Tracing the Lagrangian sub-manifold also provides rich insights into caustics (\cite{Arnold1982} and \citealt{Hidding2014}) and halo collapse \cite{Neyrinck2016}. Recently, there are attempts to improve N-body simulations (see \cite{Hahn2013}, \cite{Angulo2013a}, \cite{Angulo2013b}, \cite{Sousbie2015} and \cite{Hahn2016a}) by solving the Vlasav-Poisson equation using tessellations in the Lagrangian sub-manifold. Galaxy evolution and star formation in the context of multi streaming phenomenon are studied by \cite{Aragon-Calvo2016}.

### 2018-06-06 Draft/thesis-ku.tex
thesis update

\usepackage{mathptmx}
\renewcommand{\sfdefault}{lmss}
\renewcommand{\ttdefault}{lmtt}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{esint}   %- Commented by Nesar - check conflict with amsmath
\usepackage[authoryear]{natbib}

@@

% Set the title and author info
\title{Please Read the Abstract}
\author{Nesar Ramachandra}

\dept{Department of People who read Abstracts}
\degreetitle{Doctor of Philosophy}
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Hume Feldman}{Barbary Anthony-Twarog}{Mikhail Medvedev}{Suzanne Shontz}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

%% The following is OPTIONAL. Remove all 3 of the next 3 lines
%% to leave dates blank. If dates are included, then both dates 
%% must be included.
\@printd@testrue
\datedefended{October 02, 2018}
\dateapproved{October 03, 2018}

@@

% 2015 Paper
We report the results of the first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations
in the $\Lambda$CDM cosmology. The full dynamical state of  dark matter can be described as a three-dimensional sub-manifold
in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
(where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis.
Our major results can be summarized as follows. At the resolution of the simulation i.e. without additional
smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly 
at the filament crossings, 
and each filament is embedded in the wall like fabric of the web at the wall crossings. 
Locally, each halo or sub-halo is a peak in the number of streams field. The number of streams in the neighbouring  
filaments is higher than in the neighbouring walls. The walls are regions where number of streams
 is equal to three or a few. Voids are uniquely defined by the local condition requiring to be a single-stream flow region.
  The shells of streams 
 around haloes are  quite thin and the closest void region is typically within one and a half FOF radius 
 from the center of the halo.

% 2017 Paper_b
Mysterious dark matter constitutes about 85\% of all mass in the Universe. Clustering of dark matter plays the dominant role in the formation of all observed structures on scales from a fraction to a few hundreds of Mega-parsecs. Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web. Haloes are the highest concentrations of dark matter and host luminous galaxies. Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations. Identifying the haloes from the distribution of particles in N-body simulations is one of the problems attracting both considerable interest and efforts. We propose a novel framework for detecting potential dark matter haloes using the field unique for dark matter  -- multistream field. The multistream field emerges at the nonlinear stage of the growth of perturbations because the dark matter is collisionless. Counting the number of velocity streams in gravitational collapses supplements our knowledge of spatial clustering. We assume that the virialized haloes have convex boundaries. Closed and convex regions of the multistream field are hence isolated by imposing a positivity condition on all three eigenvalues of the Hessian estimated on the smoothed multistream field. In a single-scale analysis of high multistream field resolution and low softening length, the halo substructures with local multistream maxima are isolated as individual halo sites.

% 2017 Paper_a
Topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields. A single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than simulation resolution reveal tiny pockets of voids that are isolated by membranes of the structure. For the multistreaming excursion sets, the percolating structure is significantly  thinner than the filaments in over-density excursion approach.

Hessian eigenvalues of the multistream field are used as local geometrical indicators of dark matter structures. Single-streaming regions have most of the zero eigenvalues. Parameter-free conditions on the eigenvalues in the multistream region may be used to delineate primitive geometries with concavities corresponding to filaments, walls and haloes.

@@

% 2018 Paper
The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other. In this paper we bring twelve of these methods together and apply them to the same data set in order to understand how they compare. In general these cosmic web classifiers have been designed with different cosmological goals in mind, and to study different questions. Therefore one would not {\it a priori} expect agreement between different techniques however, many of these methods do converge on the identification of specific features. In this paper we study the agreements and disparities of the different methods. For example, each method finds that knots inhabit higher density regions than filaments, etc. and that voids have the lowest densities. For a given web environment, we find substantial overlap in the density range assigned by each web classification scheme. We also compare classifications on a halo-by-halo basis; for example, we find that 9 of 12 methods classify around a third of group-mass haloes (i.e. $M_{\rm halo}\sim10^{13.5}h^{-1}M_{\odot}$) as being in filaments. Lastly, so that any future cosmic web classification scheme can be compared to the 12 methods used here, we have made all the data used in this paper public.

@@

I would like to thank mum, dad, profs, committee members, officemates, friends, exes, uncles, aunts, grandpa, grandma, grandma's dogs, neighbors, neighbors' dogs who made this thesis possible.

@@

\bibliographystyle{apalike2}
%\bibliography{Biblio/allcites}
\bibliography{Biblio/library}

>>>>

\usepackage{mathptmx}
\renewcommand{\sfdefault}{lmss}
\renewcommand{\ttdefault}{lmtt}
\usepackage[T1]{fontenc}
% \usepackage[utf8]{inputenc}  -- Commented out due to 1.305 inputspec issue

% \usepackage[T1,T2A]{fontenc}

\usepackage{listings}
\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
\setcounter{secnumdepth}{3}
\setcounter{tocdepth}{3}
\usepackage{url}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{esint}   %- Commented by Nesar - check conflict with amsmath
\usepackage[authoryear]{natbib}

@@

% Set the title and author info
\title{Dark matter web in the Universe}
\author{Nesar Ramachandra}

\dept{Department of Physics and Astronomy}
\degreetitle{Doctor of Philosophy}
\papertype{Dissertation} %capitalization is important here
\committee{Sergei Shandarin}{Hume Feldman}{Barbary Anthony-Twarog}{Mikhail Medvedev}{Suzanne Shontz}
%AT Most 5 members allowed, last here is blank on purpose to demonstrate
%flexibility

@@

%% The following is OPTIONAL. Remove all 3 of the next 3 lines
%% to leave dates blank. If dates are included, then both dates 
%% must be included.

% \@printd@testrue
% \datedefended{July 20, 2018}
% \dateapproved{July 20, 2018}

@@

% 2015 Paper
% We report the results of the first study of the multi-stream environment of dark matter haloes in cosmological N-body simulations
% in the $\Lambda$CDM cosmology. The full dynamical state of  dark matter can be described as a three-dimensional sub-manifold
% in six-dimensional phase space - the dark matter sheet. In our study we use a Lagrangian sub-manifold ${\bf x} = {\bf x}({\bf q},t)$ 
% (where {\bf x} and {\bf q} are co-moving Eulerian and Lagrangian coordinates respectively), which 
% is dynamically  equivalent to the dark matter sheet but is more convenient for numerical analysis.
% Our major results can be summarized as follows. At the resolution of the simulation i.e. without additional
% smoothing, the cosmic web represents a hierarchical structure: each halo is embedded in the filamentary framework of the web predominantly 
% at the filament crossings, 
% and each filament is embedded in the wall like fabric of the web at the wall crossings. 
% Locally, each halo or sub-halo is a peak in the number of streams field. The number of streams in the neighbouring  
% filaments is higher than in the neighbouring walls. The walls are regions where number of streams
%  is equal to three or a few. Voids are uniquely defined by the local condition requiring to be a single-stream flow region.
%   The shells of streams 
%  around haloes are  quite thin and the closest void region is typically within one and a half FOF radius 
%  from the center of the halo.

% 2017 Paper_b

% 2017 Paper_a
% Topological connections in the single-streaming voids and multistreaming filaments and walls reveal a cosmic web structure different from traditional mass density fields. A single void structure not only percolates the multistream field in all the directions, but also occupies over 99 per cent of all the single-streaming regions. Sub-grid analyses on scales smaller than simulation resolution reveal tiny pockets of voids that are isolated by membranes of the structure. For the multistreaming excursion sets, the percolating structure is significantly  thinner than the filaments in over-density excursion approach.  
% 
% Hessian eigenvalues of the multistream field are used as local geometrical indicators of dark matter structures. Single-streaming regions have most of the zero eigenvalues. Parameter-free conditions on the eigenvalues in the multistream region may be used to delineate primitive geometries with concavities corresponding to filaments, walls and haloes.

@@

Galaxies play a role of lights illuminating these structures so they can be observed. The observations in the last several decades have unveiled opulent geometry of these structures currently known as the cosmic web. Haloes are the highest concentrations of dark matter and host luminous galaxies. Currently the most accurate modeling of dark matter haloes is achieved in cosmological N-body simulations.

% 2018 Paper
The cosmic web is one of the most striking features of the distribution of galaxies and dark matter on the largest scales in the Universe. It is composed of dense regions packed full of galaxies, long filamentary bridges, flattened sheets and vast low density voids. The study of the cosmic web has focused primarily on the identification of such features, and on understanding the environmental effects on galaxy formation and halo assembly. As such, a variety of different methods have been devised to classify the cosmic web -- depending on the data at hand, be it numerical simulations, large sky surveys or other.

@@

So many people to thank here ---

@@

\bibliographystyle{apalike2}
%\bibliography{Biblio/allcites}
\bibliography{Biblio/library3}
